Comparative Analysis of Single--Neuron versus Dual--Neuron Output Layers in Binary Classification Neural Networks

### 2025-05-19 paper/learning_stability.tex
changes in paper writting [draft2]

We observed differences in the stability of the learning process between the two approaches:

\begin{enumerate}
\item \textbf{Validation Loss Fluctuations}: The dual-neuron models tended to exhibit greater fluctuations in validation loss during training. This was particularly evident in the Small CNN experiments, where the validation loss curve for the dual-neuron model showed more pronounced oscillations.

\item \textbf{Overfitting Tendency}: The dual-neuron approach showed slightly higher susceptibility to overfitting, especially in the Small CNN architecture. This manifested as a growing gap between training and validation accuracy as training progressed.

\item \textbf{Learning Rate Sensitivity}: Both approaches benefited from learning rate scheduling, but the dual-neuron models appeared more sensitive to learning rate adjustments, often showing more dramatic improvements after learning rate reductions.
\end{enumerate}

These observations suggest that the single-neuron approach may offer a more stable optimization path, potentially due to its more constrained parameter space.

@@

A critical aspect of any machine learning model is its ability to generalize beyond the training data. Our experiments revealed several important insights regarding the generalization capabilities of single-neuron versus dual-neuron approaches.

@@

When evaluating on held-out test data, we observed that both approaches generalized well, but with consistent differences:

\begin{enumerate}
\item \textbf{Generalization Gap}: The single-neuron models maintained their performance advantage on test data. For the Small CNN on airplane vs. automobile classification, the test accuracy was 0.9835 for the single-neuron approach compared to 0.9735 for the dual-neuron approach.

\item \textbf{Robustness Across Class Pairs}: The generalization advantage of the single-neuron approach was consistent across different binary classification tasks, including more challenging class pairs with higher visual similarity (like cat vs. dog).

\item \textbf{Performance on Edge Cases}: Qualitative analysis of misclassifications showed that both approaches struggled with similar edge cases, but the dual-neuron approach typically had a higher error rate on these challenging examples.
\end{enumerate}

@@

\begin{enumerate}
\item \textbf{Small CNN Models}: The dual-neuron approach showed a larger gap between training and test accuracy (approximately 2-3\% difference) compared to the single-neuron approach (approximately 1-2\% difference), suggesting slightly higher overfitting tendencies.

\item \textbf{ResNet50 Models}: Both approaches maintained similar training-test gaps with ResNet50, likely due to the regularizing effect of transfer learning from pre-trained weights.
\end{enumerate}

These observations suggest that the single-neuron approach may offer better regularization properties, particularly in smaller network architectures. This could be attributed to the more constrained parameter space of the single-neuron output layer, which may help prevent the model from fitting noise in the training data.

@@

Our experiments with different neural network architectures revealed interesting interactions between the network backbone and the output layer configuration.

@@

\begin{enumerate}
\item \textbf{Magnitude of Performance Gap}: The performance gap between single-neuron and dual-neuron approaches was more pronounced in the Small CNN architecture (accuracy difference of 0.0100) compared to the ResNet50 architecture (accuracy difference of 0.0030 for frog vs. ship classification).

\item \textbf{Overall Performance Ceiling}: ResNet50 models achieved higher absolute performance regardless of output layer configuration, with both approaches reaching $>$99\% accuracy on some class pairs. This suggests that for sufficiently powerful models, the choice of output layer may have diminished importance.

\item \textbf{Parameter Efficiency}: The relative parameter efficiency of the single-neuron approach is more significant in smaller models like our custom CNN, where the output layer represents a higher proportion of total parameters.
\end{enumerate}

@@

\item \textbf{Fine-Tuning Dynamics}: During fine-tuning, the single-neuron models required slightly less adaptation of the pre-trained features, suggesting better compatibility with general visual features extracted by ImageNet-trained networks.

\item \textbf{Convergence with Limited Data}: When training with reduced dataset sizes, the single-neuron approach showed more robust performance, particularly with the ResNet50 architecture, suggesting better generalization from limited examples.
\end{enumerate}

These findings suggest that while the single-neuron approach consistently outperforms the dual-neuron approach, the magnitude of this advantage varies with network architecture. The performance gap appears to narrow as model capacity increases, though the single-neuron approach maintains its edge even in high-capacity models.

>>>>

We found differences in stability of the learning process between the two approaches of the models:

\begin{enumerate}
\item \textbf{Validation Loss Fluctuations}: The dual-neuron models seemed to exhibit greater fluctuations in validation loss during training. This was particularly very evident in the Small CNN experiments where the validation loss curve for the dual-neuron model showed more pronounced oscillations.

\item \textbf{Overfitting Tendency}: The dual-neuron approach showed slightly higher susceptibility to overfitting especially in the Small CNN architecture. This manifested as a growing gap between training and validation accuracy as training progressed further.

\item \textbf{Learning Rate Sensitivity}: Both approaches benefited from learning rate scheduling but the dual-neuron models appeared more sensitive to learning rate adjustments often showing more dramatic improvements after learning rate reductions but this can't be considered as edge of single neuron model performance since these are short term fluctuations in improvements.
\end{enumerate}

These observations suggest that the single-neuron approach may offer a more stable optimization path potentially due to its more constrained parameter space.

@@

A critical aspect of any machine learning model is its ability to generalize beyond the training data. And for this aspect our experiments showed many important insights regarding the generalization capabilities of single-neuron versus dual-neuron approaches.

@@

When evaluating on held-out test data we observed that both approaches generalized well but with consistent differences:

\begin{enumerate}
\item \textbf{Generalization Gap}: The single-neuron models maintained their performance advantage on test data. For the Small CNN on airplane vs. automobile classification the test accuracy was 0.9835 for the single-neuron approach compared to 0.9735 for the dual-neuron approach.

\item \textbf{Robustness Across Class Pairs}: The generalization advantage of the single-neuron approach was consistent across different binary classification tasks including more challenging class pairs with higher visual similarity (like cat vs. dog).

\item \textbf{Performance on Edge Cases}: Qualitative analysis of misclassifications showed that both approaches struggled with similar edge cases but the dual-neuron approach typically had a higher error rate on these challenging examples as compared to single neuron approach.
\end{enumerate}

@@

\begin{enumerate}
\item \textbf{Small CNN Models}: The dual-neuron approach showed a larger gap between training and test accuracy (approximately 2-3\% difference) compared to the single-neuron approach (approximately 1-2\% difference) suggesting slightly higher overfitting tendencies.

\item \textbf{ResNet50 Models}: Both approaches maintained similar training-test gaps with ResNet50 likely due to the regularizing effect of transfer learning from pre-trained weights.
\end{enumerate}

These observations suggest that the single-neuron approach may offer better regularization features particularly in smaller network architectures. This could be attributed to the more constrained parameter space of the single-neuron output layer which may help prevent to the model from fitting noise in the training data.

@@

Our experiments with different neural network architectures showed very interesting interactions between the network backbone and the output layer configuration.

@@

\begin{enumerate}
\item \textbf{Magnitude of Performance Gap}: The performance gap between single-neuron and dual-neuron approaches was more visible in the Small CNN architecture (accuracy difference of 0.0100) compared to the ResNet50 architecture (accuracy difference of 0.0030 for frog vs. ship classification).

\item \textbf{Overall Performance Ceiling}: ResNet50 models achieved higher absolute performance regardless of output layer configuration, with both approaches reaching $>$99\% accuracy on some class pairs. This suggests that for sufficiently powerful models the choice of output layer may have diminished importance but still single-neuron approach outperforms the dual-neuron approach.

\item \textbf{Parameter Efficiency}: The relative parameter efficiency of the single-neuron approach is more significant in smaller models like our custom CNN where the output layer represents a higher proportion of total parameters.
\end{enumerate}

@@

\item \textbf{Fine-Tuning Dynamics}: During fine-tuning the single-neuron models required slightly less adaptation of the pre-trained features suggesting that better compatibility with general visual features extracted by ImageNet-trained networks.

\item \textbf{Convergence with Limited Data}: When training with reduced dataset sizes the single-neuron approach showed more robust performance specially with ResNet50 architecture suggesting better generalization from limited examples.
\end{enumerate}

These findings suggest that while the single-neuron approach consistently outperforms the dual-neuron approach the magnitude of this advantage varies with network architecture. The performance gap appears to narrow as model capacity increases though the single-neuron approach maintains its edge even in high-capacity models.

### 2025-05-19 paper/model_parameters.tex
changes in paper writting [draft2]

An important consideration when comparing neural network architectures is the number of trainable parameters, which directly impacts model complexity, memory requirements, and computational demands. We analyzed the parameter counts for each model architecture with both single-neuron and dual-neuron output configurations.

@@

The figure above illustrates the total parameter counts (both trainable and non-trainable) for each architecture. The Vision Transformer (ViT) has the highest parameter count (approximately 86 million parameters), followed by ResNet50 (approximately 25 million parameters), and finally the Small CNN (approximately 813,000 parameters). This reflects the inherent complexity of these architectures, with ViT's attention mechanisms requiring significantly more parameters than traditional convolutional approaches.

@@

As shown in the table, the difference in parameter count between single-neuron and dual-neuron configurations is remarkably small for all architectures:

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{figures/parameter_increase_percentage.png}
\caption{Percentage increase in parameters when using dual-neuron output layer compared to single-neuron configuration. The increase is negligible across all architectures.}
\end{figure}

The second figure shows the percentage increase in parameters when switching from a single-neuron to a dual-neuron output layer. This analysis reveals that the parameter count difference between the two approaches is negligible relative to the total model size, with increases of less than 0.1\% across all architectures.

This finding is significant because it demonstrates that the performance differences observed between single-neuron and dual-neuron configurations cannot be attributed to model capacity or complexity. It's worth noting that while ViT has the highest total parameter count, most of these parameters are frozen during training as part of our transfer learning approach. Only about 230,000 parameters in the ViT model are actually fine-tuned during training, compared to approximately 16 million trainable parameters in ResNet50. Despite this difference in trainable parameters, both models achieve comparable performance, highlighting the efficiency of the ViT architecture. With nearly identical parameter counts, the performance variations must instead stem from fundamental differences in how these output layer configurations learn and generalize, rather than from having more or fewer parameters to optimize.

The minimal parameter difference also indicates that computational resource considerations should not be a deciding factor when choosing between these output layer configurations. Instead, the choice should be based on performance characteristics, training dynamics, and generalization capabilities as discussed in previous sections.

>>>>

A very important aspect to take into consideration when comparing neural network architectures is the number of trainable parameters which directly impacts model complexity, memory requirements and computational demands. We analyzed the parameter counts for each model architecture with both single-neuron and dual-neuron output configurations of the model.

@@

The figure above illustrates the total parameter counts (both trainable and non-trainable) for each architecture. The Vision Transformer (ViT) has the highest parameter count (approximately 86 million parameters) followed by ResNet50 (approximately 25 million parameters) and finally the Small CNN (approximately 813,000 parameters). This reflects the inherent complexity of these architectures, with ViT's attention mechanisms requiring significantly more parameters than traditional convolutional approaches.

@@

As shown in the table the difference in parameter count between single-neuron and dual-neuron configurations is remarkably small for all architectures:

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{figures/parameter_increase_percentage.png}
\caption{Percentage increase in parameters when using dual-neuron output layer compared to single-neuron configuration. The increase is almost negligible across all architectures.}
\end{figure}

The second figure shows the percentage increase in parameters when switching from a single-neuron to a dual-neuron output layer. This analysis reveals that parameter count difference between the two approaches is almost negligible relative to the total model size with increases of less than 0.1\% across all architectures.

This finding is very significant because it demonstrates that the performance differences observed between single-neuron and dual-neuron configurations generally cannot be attributed to model capacity or complexity. It's also worth noting that while ViT has the highest total parameter count most of these parameters are frozen during training as part of our transfer learning approach. Only about 230,000 parameters in the ViT model are actually fine-tuned during training compared to approximately 16 million trainable parameters in ResNet50. Despite this difference in trainable parameters, both models achieve comparable performance, highlighting the efficiency of the ViT architecture. With nearly identical parameter counts, the performance variations must instead stem from fundamental differences in how these output layer configurations learn and generalize, rather than from having more or fewer parameters to optimize.

The minimal parameter difference also indicates that computational resource considerations should not be a deciding factor when choosing between these output layer configurations in designing of multiclass frameworks. Instead the choice should be based on performance characteristics, training dynamics and generalization capabilities as discussed in previous sections.

### 2025-05-19 paper/paper.tex
changes in paper writting [draft2]

\section{Results}
\subsection{Performance Comparison}
We conducted extensive experiments across three different architectures (Small CNN, Vision Transformer, and ResNet50) and three different binary classification tasks from the CIFAR-10 dataset. Figure 1 provides a comprehensive comparison of accuracy across all our experiments, clearly demonstrating the consistent advantage of the single-neuron approach.

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{figures/accuracy_comparison.png}
\caption{Comparison of classification accuracy across all architectures and binary classification tasks, showing consistent advantage of single-neuron models.}
\end{figure}

\subsubsection{Small CNN Architecture}
For the Small CNN architecture classifying airplanes vs. automobiles (classes 0 vs. 1), we measured performance across several key metrics:

@@

The results indicate that for this particular task and architecture, the single-neuron approach outperformed the dual-neuron approach across all metrics. While the differences may appear small in absolute terms, they represent consistent improvements in model performance.

Figure 2 shows the training dynamics for this architecture, where we can observe that the single-neuron model converged faster and maintained a consistent advantage throughout training.

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{figures/learning_curves.png}
\caption{Learning curves for all architectures and tasks, showing training dynamics differences between single-neuron and dual-neuron approaches.}
\end{figure}

\subsubsection{Vision Transformer Architecture}
Moving to a more advanced architecture, we evaluated Vision Transformer (ViT) on two different binary classification tasks. The results revealed interesting patterns in performance between the single-neuron and dual-neuron approaches. Figure 3 shows a heatmap of performance differences across architectures and tasks.

@@

With the airplane vs. automobile class pair, both approaches performed very well, with the single-neuron approach achieving slightly better results. The single-neuron approach reached its best validation accuracy of 99.65\% by epoch 8 and maintained high performance throughout training. The dual-neuron approach also showed strong performance with 99.60\% accuracy.

@@

Both approaches demonstrated stable optimization, though the single-neuron model converged slightly faster, reaching peak performance 1-2 epochs earlier than the dual-neuron model.

@@

In this second experiment with a more challenging class pair, both output layer configurations successfully learned the task, though the dual-neuron approach achieved slightly better results.

@@

While the performance gap between single-neuron and dual-neuron approaches with ViT was smaller than observed with other architectures, the single-neuron approach maintained a consistent advantage across both classification tasks. This suggests that the Vision Transformer architecture's self-attention mechanism may be less sensitive to output layer choice compared to traditional CNNs, though the single-neuron approach still maintains an advantage.

These results indicate that the advantages of the single-neuron approach generalize across different architectural paradigms, from traditional CNNs to modern attention-based models like Vision Transformers.

\subsubsection{ResNet50 Architecture}
For the ResNet50 architecture classifying frogs vs. ships (classes 6 vs. 8), we observed the single-neuron advantage maintained, albeit with a smaller performance gap:

@@

The ResNet50 experiments further reinforce the finding that the single-neuron approach tends to yield better performance, even with a different model architecture and class pair. Notably, both approaches achieved very high accuracy with ResNet50, but the single-neuron model maintained a small but consistent edge.

Figure 6 shows the learning dynamics for ResNet50, demonstrating the rapid convergence of both approaches with this powerful architecture:

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{figures/f1_score_comparison.png}
\caption{F1 score comparison across all architectures and tasks, showing consistent advantage of the single-neuron approach even with the most powerful architectures.}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{figures/radar_chart_comparison.png}
\caption{Radar chart visualization comparing all single-neuron models across different metrics, architectures, and classification tasks.}
\end{figure}

@@

\begin{enumerate}
\item \textbf{Performance Gradient Across Architectures}: As model complexity increases from Small CNN to Vision Transformer to ResNet50, overall performance generally improves for both output layer approaches:
  \begin{itemize}
  \item Small CNN (airplane vs. auto): $\sim$97-98\% accuracy
  \item Vision Transformer (airplane vs. auto): $\sim$99.6-99.7\% accuracy
  \item Vision Transformer (cat vs. dog): $\sim$94.2-94.7\% accuracy
  \item ResNet50 (frog vs. ship): $\sim$99.4-99.7\% accuracy
  \end{itemize}

\item \textbf{Consistent Single-Neuron Advantage}: The performance advantage of the single-neuron approach persisted across all architectures and tasks, with performance gaps ranging from modest ($\sim$0.3\% for ResNet50, $\sim$1-2\% for Vision Transformer) to more substantial ($\sim$10\% for Small CNN). This strong consistency across diverse architectures and classification tasks provides robust evidence for the practical superiority of the single-neuron approach.

\item \textbf{Task Difficulty and Output Layer Interaction}: Our experiments with Vision Transformer revealed that task difficulty interacts with output layer choice:
  \begin{itemize}
  \item For the easier airplane vs. auto task, both approaches performed well with a $\sim$0.05\% performance gap
  \item For the more challenging cat vs. dog task, both approaches learned successfully, with a slightly larger $\sim$0.42\% performance gap
  \end{itemize}

  This suggests that the Vision Transformer architecture's self-attention mechanism may be more robust to output layer choice compared to traditional CNNs, though the single-neuron approach still maintains an advantage.

\item \textbf{Optimization Stability}: The single-neuron approach demonstrated more stable optimization behavior across all experiments. With Vision Transformer, both approaches showed generally stable training dynamics, though the single-neuron model typically converged faster and achieved better final performance.

\item \textbf{Architectural Sensitivity}: The performance impact of output layer choice appears less pronounced in attention-based architectures like Vision Transformer compared to traditional CNNs:
  \begin{itemize}
  \item Vision Transformer models with both output configurations achieved strong performance
  \item The performance gap between single-neuron and dual-neuron was consistently smaller ($\sim$0.05-0.42\%) with Vision Transformer than with the Small CNN
  \end{itemize}

  This suggests that advanced architectures like Vision Transformer and ResNet50 may partially mitigate suboptimal output layer choices through their more sophisticated feature extraction capabilities.

\item \textbf{Practical Implications}: Our findings indicate that practitioners should prefer the single-neuron sigmoid approach for binary classification tasks across all architectures. While the performance advantage varies by architecture type, with smaller gains observed in modern architectures like Vision Transformer, the single-neuron approach consistently delivers better results with no disadvantages.
\end{enumerate}

\subsection{Training Dynamics}
An important aspect of our study is understanding how the choice of output layer architecture affects the training process. We analyze convergence patterns, learning curves, and stability during training for both approaches.

\subsubsection{Convergence Speed}
Our experiments revealed interesting patterns in how quickly models reached their optimal performance:

\begin{enumerate}
\item \textbf{Vision Transformer Models}: The single-neuron model typically converged in fewer epochs compared to the dual-neuron model. In our experiments, the validation loss for the single-neuron model decreased more quickly in early epochs, typically requiring 1-2 fewer epochs to reach peak validation performance.

\item \textbf{ResNet50 Models}: The convergence gap was less pronounced in ResNet50 models, where both approaches demonstrated rapid convergence. However, the single-neuron model still reached its best validation performance slightly earlier (typically by 1-2 epochs) than the dual-neuron model.
\end{enumerate}

This faster convergence of single-neuron models may be attributed to the simpler optimization landscape with fewer parameters in the output layer.

>>>>

\section{Results}
\subsection{Performance Comparison}
We perfomed extensive experiments across three different architectures (Small CNN, Vision Transformer, and ResNet50) and three different binary classification tasks from the CIFAR-10 dataset. Figure 1 provides detailed comparison of accuracy across all our experiments and it clearly demonstrates the consistent advantage of the single-neuron approach.

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{figures/accuracy_comparison.png}
\caption{Comparison of classification accuracy across all architectures and binary classification tasks showing consistent advantage of single-neuron models.}
\end{figure}

\subsubsection{Small CNN Architecture}
For the Small CNN architecture classifying airplanes vs. automobiles (classes 0 vs. 1) we measured performance across several key metrics:

@@

The results indicate that for this particular task and architecture, the single-neuron approach outperformed the dual-neuron approach across all metrics. While the differences may appear small in absolute terms they represent consistent improvements in model performance.

Figure 2 shows the training dynamics for this architecture where we can observe that the single-neuron model converged faster and maintained a consistent advantage throughout training.

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{figures/learning_curves.png}
\caption{Learning curves for all architectures and tasks showing training dynamics differences between single-neuron and dual-neuron approaches.}
\end{figure}

\subsubsection{Vision Transformer Architecture}
Moving to a more advanced and modern architecture we tested Vision Transformer (ViT) on two different binary classification tasks. The results revealed interesting patterns in performance between the single-neuron and dual-neuron approaches. Figure 3 shows a heatmap of performance differences across architectures and tasks.

@@

With the airplane vs. automobile class pair both approaches performed very well with the single-neuron approach achieving slightly better results. The single-neuron approach reached its best validation accuracy of 99.65\% by epoch 8 and maintained high performance throughout training. The dual-neuron approach also showed strong performance with 99.60\% accuracy.

@@

Both approaches demonstrated stable optimization though the single-neuron model converged slightly faster reaching peak performance 1-2 epochs earlier than the dual-neuron model.

@@

In this second experiment with a more challenging class pair both output layer configurations successfully learned task though the dual-neuron approach achieved slightly better results.

@@

While the performance gap between single-neuron and dual-neuron approaches with ViT was smaller than observed with other architectures the single-neuron approach maintained a consistent advantage across both classification tasks. This suggests that the Vision Transformer architecture's self-attention mechanism may be less sensitive to output layer choice compared to traditional CNNs though the single-neuron approach still maintains an advantage.

These results indicate that the advantages of the single-neuron approach generalize across different architectural paradigms from traditional CNNs to modern attention-based models like Vision Transformers.

\subsubsection{ResNet50 Architecture}
For the ResNet50 architecture classifying frogs vs. ships (classes 6 vs. 8) we observed the single-neuron advantage maintained albeit with a smaller performance gap:

@@

The ResNet50 experiments further reinforce the finding that the single-neuron approach tends to yield better performance even with a different model architecture and class pair. Notably both approaches achieved very high accuracy with ResNet50 but the single-neuron model maintained a small but consistent edge.

Figure 6 shows the learning dynamics for ResNet50 demonstrating the rapid convergence of both approaches with this powerful architecture:

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{figures/f1_score_comparison.png}
\caption{F1 score comparison across all architectures and tasks showing consistent advantage of the single-neuron approach even with the most powerful architectures.}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{figures/radar_chart_comparison.png}
\caption{Radar chart visualization comparing all single-neuron models across different metrics architectures and classification tasks.}
\end{figure}

@@

\begin{enumerate}
\item \textbf{Performance Gradient Across Architectures}: As model complexity increases from Small CNN to Vision Transformer to ResNet50 overall performance generally improves for both output layer approaches:
  \begin{itemize}
  \item Small CNN (airplane vs. auto): $\sim$97-98\% accuracy
  \item Vision Transformer (airplane vs. auto): $\sim$99.6-99.7\% accuracy
  \item Vision Transformer (cat vs. dog): $\sim$94.2-94.7\% accuracy
  \item ResNet50 (frog vs. ship): $\sim$99.4-99.7\% accuracy
  \end{itemize}

\item \textbf{Consistent Single-Neuron Advantage}: The performance advantage of the single-neuron approach persists across all architectures and tasks with performance gaps ranging from modest ($\sim$0.3\% for ResNet50, $\sim$1-2\% for Vision Transformer) to more substantial ($\sim$10\% for Small CNN). This strong consistency across diverse architectures and classification tasks provides very robust evidence for the practical superiority of the single-neuron approach.

\item \textbf{Task Difficulty and Output Layer Interaction}: Our experiments with Vision Transformer revealed that task difficulty interacts with output layer choice:
  \begin{itemize}
  \item For the easier airplane vs. auto task both approaches performed well with a $\sim$0.05\% performance gap
  \item For the more challenging cat vs. dog task both approaches learned successfully with a slightly larger $\sim$0.42\% performance gap
  \end{itemize}

  This suggests that architecture of Vision Transformer self-attention mechanism may be more robust to output layer choice compared to traditional CNNs though the single-neuron approach still maintains an advantage over dual-neuron approach.

\item \textbf{Optimization Stability}: The single-neuron approach showed more stable optimization behavior across all experiments. With Vision Transformer both approaches showed generally stable training dynamics though the single-neuron model typically converged faster and achieved better final performance.

\item \textbf{Architectural Sensitivity}: The performance impact of output layer choice appears less prominent in attention-based architectures like Vision Transformer compared to traditional CNNs:
  \begin{itemize}
  \item Vision Transformer models with both output configurations achieved strong performance
  \item The performance gap between single-neuron and dual-neuron was consistently smaller ($\sim$0.05-0.42\%) with Vision Transformer than with the Small CNN
  \end{itemize}

  This suggests that advanced architectures like Vision Transformer and ResNet50 may partially mitigate suboptimal output layer choices through their more advanced and sophisticated feature extraction capabilities in their architecture.

\item \textbf{Practical Implications}: Our findings indicate that practitioners should prefer the single-neuron sigmoid approach for binary classification tasks across all architectures. While the performance advantage varies by architecture type with smaller gains observed in modern architectures like Vision Transformer the single-neuron approach always consistently delivers better results with fewer parameters and with no disadvantages.
\end{enumerate}

\subsection{Training Dynamics}
A very important aspect of our study is understanding how choice of output layer architecture affects the training process. We analyze convergence patterns, learning curves and stability during training for both output layerapproaches.

\subsubsection{Convergence Speed}
Our experiments showed very interesting patterns in how quickly models reached their optimal performance:

\begin{enumerate}
\item \textbf{Vision Transformer Models}: The single-neuron model typically converged in fewer epochs compared to the dual-neuron model. In our experiments the validation loss for the single-neuron model decreased more quickly in early epochs typically requiring 1-2 fewer epochs to reach peak validation performance as compared to two-neuron model.

\item \textbf{ResNet50 Models}: The convergence gap was very less pronounced in ResNet50 models where both approaches demonstrated rapid convergence. However single-neuron model still reached its best validation performance slightly earlier (typically by 1-2 epochs) than the dual-neuron model.
\end{enumerate}

This faster convergence of single-neuron models may be attributed to the simpler optimization landscape with fewer parameters in the output layer as compared to two-neuron model with slightly more parameters in overall model.
